Parsing Postfix log files

### 2008-10-31 doc/logparser-acronyms.tex
Finished corrections; 112 XXX notes.

\newacronym{PLP}{Postfix Log Parser}{description={
    Postfix Log Parser is the implementation of the algorithm described in
    this paper.
}}

@@

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and receives mail via
    SMTP\backslashat{}.
}}

@@

\newacronym{SPF}{Sender Policy Framework}{description={
    SPF is introduced in \sectionref{spf introduction} and explained fully
    in~\cite{openspf,Wikipedia-spf}.
}}

\newacronym{ISP}{Internet Service Provider}{description={
    An ISP is a company that sells Internet access to consumers.
}}

\newacronym{LMA}{Log Mail Analyzer}{description={
    One of the other Postfix log parsers reviewed and the only example of
    published prior art~\cite{log-mail-analyser} found that specifically
    deals with parsing Postfix logs.
}}

@@

\newacronym{DNS}{Domain Name System}{description={
    The DNS converts between hostnames (www.example.com) and IP addresses
    (10.1.2.3).
}}

@@

\newacronym{regex}{Regular Expression}{description={
    Regular expressions are a concise and powerful method of matching
    patterns against text that are explained in detail
    in~\cite{Wikipedia-regex,perlre}.
}}

% I think this is better than having both regex and regexes in the acronym
% list.  Likewise for pids.
\newcommand{\regexes}{\regex{}es}
\newcommand{\pids}{\pid{}s}

@@

\newacronym{IP}{Internet Protocol}{description={
    The Internet Protocol~\cite{Wikipedia-ip} is the protocol used to
    communicate between computers on the Internet.  An IP address is a
    unique address assigned to a computer on the Internet, allowing it to
    communicate with other computers on the Internet.  For more information
    see~\cite{Wikipedia-ip-address}.
}}

\newacronym{PDF}{Portable Document Format}{description={
    PDF is the file format created by Adobe Systems in 1993 for document
    exchange.  It aims to be device independent, so documents should look
    the same whether viewed on screen or printed.
}}

\newacronym{LCD}{Lowest Common Denominator}{description={
    LCD is a mathematical term that is used figuratively to mean the least
    useful member of a set of alternatives.
}}

>>>>

\newacronym{PLP}{Postfix Log Parser}{description={
    Postfix Log Parser is the parser implementing the algorithm described
    in this thesis.
}}

@@

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and receives mail via
    SMTP\backslashat{}.  XXX DO I NEED THIS\backslashat{}?
}}

@@

\newacronym{LMA}{Log Mail Analyzer}{description={
    One of the other Postfix log parsers reviewed and the only example of
    prior published paper~\cite{log-mail-analyser} found that specifically
    deals with parsing Postfix log files.
}}

@@

@@

% I think this is better than having both regex and regexes in the acronym
% list.  Likewise for pids.
\newcommand{\pids}{\pid{}s}

@@

\newacronym{IP}{Internet Protocol}{description={
    The Internet Protocol is the protocol used to communicate between
    computers on the Internet.  An IP address is a unique address assigned
    to a computer on the Internet, allowing it to communicate with other
    computers on the Internet.
}}

### 2008-10-31 doc/logparser-architecture.tex
First half of the corrections to section 5.
Fixed all overfull hbox warnings, and made some small improvements in the process.
Finished corrections; 112 XXX notes.

XXX REWRITE TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.

XXX NEEDS AN INTRODUCTION ONCE CONTENT HAS SETTLED\@.

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a \regex{} to match the log lines is
trivial in comparison to understanding a program's entire parsing
algorithm, identifying the correct location to change and making the
appropriate changes.  Bear in mind that the changes must be made without
adversely affecting existing parsing, particularly as there may be edge
cases that are not immediately obvious ---
\sectionref{yet-more-aborted-delivery-attempts} describes a complication
that occurs only four times in \numberOFlogFILES{} log files tested.
Requiring changes to the parser's code also complicates upgrades, as the
changes must be preserved during the upgrade, and may clash with changes
made by the developer (\sectionref{adding new actions} describes safely
adding new actions).  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare event that the new actions require support from
other sections of the code will more extensive changes be required.

@@

@@

    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
        similar to: \textbf{select regex from rules order by regex;}

@@

>>>>

XXX REWRITE\@: MERGE CONTENT FROM ELSEWHERE IN THE THESIS AND PAPER\@.

XXX NEEDS AN INTRODUCTION\@.

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to match the log lines is
trivial in comparison to understanding a program's entire parsing
algorithm, identifying the correct location to change and making the
appropriate changes.  Bear in mind that the changes must be made without
adversely affecting existing parsing, particularly as there may be edge
cases that are not immediately obvious ---
\sectionref{yet-more-aborted-delivery-attempts} describes a complication
that occurs only four times in \numberOFlogFILES{} log files tested.
Requiring changes to the parser's code also complicates upgrades, as the
changes must be preserved during the upgrade, and may clash with changes
made by the developer (\sectionref{adding new actions} describes safely
adding new actions).  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare event that the new actions require support from
other sections of the code will more extensive changes be required.

@@

\label{rule in architecture}

@@

    \item Sort by regex and visually inspect the list, e.g.\ with \SQL{}
        similar to: \textbf{select regex from rules order by regex;}

@@

\label{actions in architecture}

### 2008-10-31 doc/logparser-background.tex
Finished corrections; 112 XXX notes.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible as described
in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.} The
meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing
exceptions to more general rules.  When the result is the name of another
restriction Postfix will evaluate that restriction, allowing restrictions
to be chosen based on the client \IP{} address, client hostname, HELO
hostname, sender address, or recipient address.\footnote{E.g.\ the
administrator may require that all clients on the local network have valid
\DNS{} entries, to prevent people sending mail from unknown machines.}  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.  Postfix tries to protect the administrator in
as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
--- the administrator must be cognisant of the effects of their
configuration changes.  This is similar to one of UNIX's design
philosophies: \textit{``UNIX was not designed to stop its users from doing
stupid things, as that would also stop them from doing clever
things''\/}~\cite{unix-philosophy}.

@@

    \item Checking \SPF{} records~\cite{openspf,Wikipedia-spf}.
        \SPF{}\label{spf introduction} records specify which mail servers
        are allowed to send mail claiming to be from a particular domain.
        The intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter}, and
        joe~jobs~\cite{Wikipedia-joe-job}.  There has been considerable
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}, e.g.\ forwarding mail from one company or
        university to another when a user moves.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote \IP{}
        address)}\newline{} has not been seen before; on second and
        subsequent delivery attempts from that triple the mail will be
        accepted.  This blocks spam because maintaining a list of failed
        addresses and retrying after a temporary failure is uneconomical
        for a spam sender, but a legitimate mail server must retry
        deliveries that temporarily failed.  Sadly spam senders are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used in testing the parser are shown
        in \tableref{Summary of rejections}.  Greylisting is obviously
        worth using, at least at the moment, particularly when you consider
        Greylisting's position as the final restriction that a mail must
        overcome:\footnote{Greylisting is the final restriction a mail must
        overcome in the configuration used on the mail server the log files
        were obtained from; an administrator is free to use Greylisting
        wherever in the restriction list they feel is most appropriate for
        their mail system.} Greylisting only takes effect for mails that
        have passed every other restriction.  Some problems may be
        encountered when using greylisting: some servers fail to retry;
        legitimate mail may be delayed, particularly when coming from a
        pool of servers.

>>>>

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible as described
in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.} The
meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing
exceptions to more general rules.  When the result is the name of another
restriction Postfix will evaluate that restriction, allowing restrictions
to be chosen based on the client \IP{} address, client hostname, HELO
hostname, sender address, or recipient address.\footnote{E.g.\ the
administrator may require that all clients on the local network have valid
DNS entries, to prevent people sending mail from unknown machines.}  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.  Postfix tries to protect the administrator in
as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
--- the administrator must be cognisant of the effects of their
configuration changes.  This is similar to one of UNIX's design
philosophies: \textit{``UNIX was not designed to stop its users from doing
stupid things, as that would also stop them from doing clever
things''\/}~\cite{unix-philosophy}.

@@

    \item Sender Policy Framework (SPF)~\cite{openspf}.  SPF\label{spf
        introduction} records specify which mail servers are allowed to
        send mail claiming to be from a particular domain.  The intention
        is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter}, and
        joe~jobs~\cite{Wikipedia-joe-job}.  There has been considerable
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}, e.g.\ forwarding mail from one company or
        university to another when a user moves.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote \IP{}
        address)}\newline{} has not been seen before; on second and
        subsequent delivery attempts from that triple the mail will be
        accepted.  This blocks spam because maintaining a list of failed
        addresses and retrying after a temporary failure is uneconomical
        for a spam sender, but a legitimate mail server must retry
        deliveries that temporarily failed.  Sadly spam senders are using
        increasingly complex and well written programs to distribute spam,
        frequently using an ISP provided \SMTP{} server from a compromised
        machine on the ISP's network.  Greylisting will slowly become less
        useful, but it does block a large percentage of spam mail at the
        moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used in testing the parser are shown
        in \tableref{Summary of rejections}.  Greylisting is obviously
        worth using, at least at the moment, particularly when you consider
        Greylisting's position as the final restriction that a mail must
        overcome:\footnote{Greylisting is the final restriction a mail must
        overcome in the configuration used on the mail server the log files
        were obtained from; an administrator is free to use Greylisting
        wherever in the restriction list they feel is most appropriate for
        their mail system.} Greylisting only takes effect for mails that
        have passed every other restriction.  Some problems may be
        encountered when using greylisting: some servers fail to retry;
        legitimate mail may be delayed, particularly when coming from a
        pool of servers.

### 2008-10-31 doc/logparser-conclusion.tex
Finished corrections; 112 XXX notes.

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of \regexes{} from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules}).

@@

The division of the parser into rules, actions, and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (the combination of \texttt{lex} and
\texttt{yacc}~\cite{lex-and-yacc} being an obvious example), the parts are
usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  The actions and
framework are not as completely separated, as the actions depend on
services provided by the framework; however the actions and framework are
not tightly integrated: it would be possible, with some work, to completely
separate the two.  XXX I SHOULD DO THAT\@.

% XXX THIS SHOULD PROBABLY BE REMOVED\@.
%The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
%behaviour}) exhibited by the rules and actions is also interesting, and is
%discussed after the flow chart (\sectionref{flow-chart}) and explanation of
%the paths through the parser.  This emergent behaviour greatly eases the
%process of adding new actions (\sectionref{adding new actions}), as the
%actions do not have to be inserted into an explicit flow of control; new
%actions will naturally find their place in the paths through the parser.

>>>>

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of regexes from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules}).

@@

The division of the parser into rules, actions, and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (the combination of \texttt{lex} and
\texttt{yacc} XXX BOOK REF FOR LEX AND YACC being a common example), the
parts are usually quite internally interdependent, and will be combined
into a complete parser by the compilation process; in contrast
\parsername{} keeps the rules and actions separate until the parser runs.
The actions and framework are not as completely separated, as the actions
depend on services provided by the framework; however the actions and
framework are not tightly integrated: it would be possible, with some work,
to completely separate the two.  XXX I SHOULD DO THAT\@.

% XXX THIS SHOULD PROBABLY BE REMOVED\@.
%The emergent behaviour (\sectionref{Emergent
%behaviour}) exhibited by the rules and actions is also interesting, and is
%discussed after the flow chart (\sectionref{flow-chart}) and explanation of
%the paths through the parser.  This emergent behaviour greatly eases the
%process of adding new actions (\sectionref{adding new actions}), as the
%actions do not have to be inserted into an explicit flow of control; new
%actions will naturally find their place in the paths through the parser.

### 2008-10-31 doc/logparser-glossary.tex
Finished corrections; 112 XXX notes.

\glossary{name={postsuper},description={
    Maintenance task such as deleting mails from the queue, putting mail on
    hold (no further delivery attempts will be made until it is released
    from hold, also by postsuper), and consistency checking of the mail
    queue.  The definitive documentation is available at
    \url{http://www.postfix.org/postsuper.1.html}.
}}

@@

\glossary{name={sendmail},description={
    Postfix provides a command that is compatible with the Sendmail
    (\url{http://www.sendmail.org/}) mail submission program that all Unix
    commands that send mail depend on; Postfix sendmail executes postdrop
    to place a new mail in the queue.  The definitive documentation is
    \url{http://www.postfix.org/sendmail.1.html}.
}}

@@

\glossary{name={queueid},description={
    Each mail in Postfix's queue is assigned a queueid to uniquely identify
    it.  Queueids are assigned from a limited pool, so although they are
    guaranteed to be unique for the lifetime of the mail, given sufficient
    time they will be reused.
}}

\glossary{name={IPv4},description={
    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
    of the Internet Protocol used to interconnect computers on the
    Internet.  It is the first widely deployed version of IP, and has been
    in use for over 25 years.
}}

\glossary{name={IPv6},description={
    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
    of the Internet Protocol used to interconnect computers on the
    Internet.  It is the successor to IPv4, bringing with it a greatly
    expanded address space, allowing many more computers to use the
    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
    come as existing networks transition from the former to the latter.
}}

@@

\glossary{name={Solaris},description={
    Solaris is a Unix-derived Operating System produced by Sun Microsystems
    (\url{http://www.sun.com/software/solaris/}).
}}

@@

\glossary{name={mail loop},description={
    Sometimes mail set to one address must be delivered to a different
    address instead, e.g.\ because a person has changed jobs.  A mail loop
    occurs when the recipient addresses constitute a cyclic directed graph;
    the simplest example is when mail for \texttt{foo\at{}example.net} is
    delivered to \texttt{bar\at{}example.com}, and mail for
    \texttt{bar\at{}example.com} is delivered to
    \texttt{bar\at{}example.com}.
}}

>>>>

\glossary{name={postsuper},description={
    Maintenance tasks such as deleting mails from the queue, putting mail
    on hold and later releasing it (no further delivery attempts will be
    made until it is released), and consistency checking of the mail queue.
    The definitive documentation is available at
    \url{http://www.postfix.org/postsuper.1.html}.
}}

@@

\glossary{name={sendmail},description={
    Postfix provides a command that is compatible with the Sendmail
    (\url{http://www.sendmail.org/}) mail submission program that all Unix
    commands that send mail depend on; it executes \daemon{postdrop} to
    place a new mail in the queue.  The definitive documentation is
    \url{http://www.postfix.org/sendmail.1.html}.
}}

@@

\glossary{name={queueid},description={
    Each mail in Postfix's queue is assigned a queueid to uniquely identify
    it.  Queueids are assigned from a limited pool, so although they are
    guaranteed to be unique for the lifetime of the mail, they may be
    reused later.
}}

\glossary{name={IPv4},description={
    Internet Protocol version 4 is the fourth version of the Internet
    Protocol used to interconnect computers on the Internet.  It is the
    first widely deployed version of IP, and has been in use for over 25
    years.
}}

\glossary{name={IPv6},description={
    Internet Protocol version 6 is the latest version of the Internet
    Protocol used to interconnect computers on the Internet.  It is the
    successor to IPv4, bringing with it a greatly expanded address space,
    allowing many more computers to use the Internet simultaneously.  IPv4
    and IPv6 will coexist for many years to come as existing networks
    transition from the former to the latter.
}}

@@

@@

\glossary{name={mail loop},description={
    Sometimes mail sent to one address must be forwarded to a different
    address instead, e.g.\ because a person has changed jobs.  A mail loop
    occurs when the chain of recipient addresses constitute a cyclic
    directed graph; the simplest example is when mail for
    \texttt{alice\at{}example.net} is delivered to
    \texttt{bob\at{}example.com}, and mail for \texttt{bob\at{}example.com}
    is delivered to \texttt{alice\at{}example.net}.
}}

### 2008-10-31 doc/logparser-implementation.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).
Finished corrections; 112 XXX notes.

            \item [PROCESSING] \daemon{cleanup} is processing a mail
                (see~\cite{postfix-cleanup} for details).

@@

    \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
        EXTRACTION IS FINISHED\@.  The \regex{} to match the log line
        against.  The \regex{} will first have several keywords expanded:
        this simplifies reading and writing rules; avoids needless
        repetition of complex \regex{} components; allows the components to
        be corrected and/or improved in one location; and makes each
        \regex{} largely self-documenting.  For efficiency the keywords are
        expanded and every rule's \regex{} is compiled before attempting to
        parse the log file --- otherwise each \regex{} would be recompiled
        each time it was used, resulting in a large, data dependent
        slowdown.  Efficiency concerns are discussed in \sectionref{parser
        efficiency}, with caching compiled \regexes{} covered in
        \sectionref{Caching compiled regexes}.

@@

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via \DNS{}\@.

@@

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via \DNS{}\@.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address because the domain in the sender address does
not have an A, AAAA, or MX \DNS{} entry, i.e.\ mail could not be delivered
to the sender's address (for full details
see~\cite{reject-unknown-sender-domain}).

@@

    \item [program and regex:] If the program in the rule equals the
        program in the log line, a match using the rule's \regex{} will be
        attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

    \item [action:] The action to be executed if the \regex{} matches
        successfully.  See \sectionref{actions-in-detail} for details of
        the actions implemented by the parser, \sectionref{actions in
        architecture} for the role of actions in the parser architecture.

@@

The core of the \SLCT{} algorithm is quite simple: log lines are generally
created by substituting variable words into a fixed pattern, and analysis
of the frequency with which each word occurs can be used to determine
whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a
\regex{} to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        that the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        \regexes{}; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each log line, incrementing a counter for each hash.  The
        counters will be used in the next pass to filter out words: if the
        word's hash does not have a high frequency, the word itself cannot
        have a high frequency, so there is no need to maintain a counter
        for it, reducing the number of counters required and thus the
        program's memory consumption.

@@

    \item [Classify words based on their frequency]  The frequency of every
        tuple \texttt{(word, word's position within the log line)} is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point)
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be misclassified as a fixed term, but that should be noticed
        by the user when reviewing the new \regexes{}.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.  XXX SHOULD I USE \verb![\S]+! INSTEAD\@?

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the log line, and a counter is maintained for each
        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
        resulting duplicate \regexes{} are combined, and their counters
        added together.  If a \regex{}'s counter is lower than the
        threshold supplied by the user the \regex{} is discarded; this
        second threshold is independent of the threshold used to
        differentiate between fixed and variable words, but once again 1\%
        of log lines is a good starting point.  The new \regexes{} are
        displayed for the user to add to the database, either as new rules
        or merged into the \regexes{} of existing rules; the counter for
        each \regex{} is also displayed, giving the user an indication of
        how many of the log lines that \regex{} should match.  Discarding
        \regexes{} will result in some of the log lines not being matched;
        when the rules have been augmented with the new \regexes{}, the
        original log files should be parsed again, and any remaining
        unparsed log lines used as input to this utility.

@@

A second utility is also provided that reads a list of new \regexes{} and
the input given to the first utility.  It tries to match each input log
line against each \regex{}, counting the number of log lines that match
each \regex{}, warning the user if an input log line is matched by more
than one \regex{}, and additionally warning if an input log line is not
matched by any \regex{}.  It displays a summary of how many input log lines
each \regex{} matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the \regexes{}
produced by the first utility are correctly matching the input log lines
they are based upon.  A future version of this utility will also group
input log lines by \regex{}, so the user can tweak the \regexes{} if
required.

These utilities are not intended to create perfect \regexes{}, but they
greatly reduce the effort required to parse new or different log lines.

@@

The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
that is far more complicated than the behaviour explicitly encoded in the
algorithm.  The top-level parser could be written in pseudo-code like so
(with indentation level indicating flow control):

@@

    \item [matches] The fields in the line captured by the rule's \regex{}.

@@

    \item [SAVE\_DATA] Find the correct mail based on the queueid in
        the log line, and save the data extracted by the \regex{} to it.

@@

    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
        signal~\cite{Wikipedia-unix-signals}; the active connection for
        that \daemon{smtpd} must be cleaned up, entered in the database,
        and deleted from the state tables.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations --- after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur
        in the intervening hours: \DNS{} timeouts, timeouts reading data
        from the client, etc\@.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

@@

This section presented the core of the parser, starting with a very high
level view and the initial complications that arose.  A flow chart showing
the paths a mail may take through the nascent simplified algorithm was
provided, followed by an explanation of those paths, and a discussion of
the parser's emergent behaviour --- the data from the log files creates the
paths in the flow chart, they are not specified anywhere in the parser.
The framework which holds the parser together was covered next, after which
came a description of the current actions provided by the parser, and the
algorithm for analysing unparsed log lines to create \regexes{} for new
rules.  Detecting, diagnosing and defeating complications forms the largest
single portion of this section, mirroring the development of the parser.
The complications are described in the order they were overcome, with
subsequent problems affecting fewer mails (often by an order or magnitude),
though the time required to solve problems increased with each successive
problem.

>>>>

            \item [PROCESSING] \daemon{cleanup} is processing a mail.
                XXX ADD MORE DETAIL HERE

@@

    \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
        EXTRACTION IS FINISHED\@.  The regex to match the log line
        against.  The regex will first have several keywords expanded:
        this simplifies reading and writing rules; avoids needless
        repetition of complex regex components; allows the components to
        be corrected and/or improved in one location; and makes each
        regex largely self-documenting.  For efficiency the keywords are
        expanded and every rule's regex is compiled before attempting to
        parse the log file --- otherwise each regex would be recompiled
        each time it was used, resulting in a large, data dependent
        slowdown.  Efficiency concerns are discussed in \sectionref{parser
        efficiency}, with caching compiled regexes covered in
        \sectionref{Caching compiled regexes}.

@@

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via DNS\@.

@@

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via DNS\@.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address because the domain in the sender address does
not have an A, AAAA, or MX DNS entry, i.e.\ mail could not be delivered
to the sender's address (for full details
see~\cite{reject-unknown-sender-domain}).

@@

    \item [program and regex:] If the program in the rule equals the
        program in the log line, a match using the rule's regex will be
        attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

    \item [action:] The action to be executed if the regex matches
        successfully.  See \sectionref{actions-in-detail} for details of
        the actions implemented by the parser, \sectionref{actions in
        architecture} for the role of actions in the parser architecture.

@@

The core of the \SLCT{} algorithm is quite simple: log lines are generally
created by substituting variable words into a fixed pattern, and analysis
of the frequency with which each word occurs can be used to determine
whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a
regex to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with regex keywords
        that the parser will expand when it loads the rule (see the
        regex entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        regexes; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash~\cite{hash-functions} from a small range of values for each word
        of each log line, incrementing a counter for each hash.  The
        counters will be used in the next pass to filter out words: if the
        word's hash does not have a high frequency, the word itself cannot
        have a high frequency, so there is no need to maintain a counter
        for it, reducing the number of counters required and thus the
        program's memory consumption.

@@

    \item [Classify words based on their frequency]  The frequency of every
        tuple \texttt{(word, word's position within the log line)} is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point)
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be misclassified as a fixed term, but that should be noticed
        by the user when reviewing the new regexes.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.  XXX SHOULD I USE \verb![\S]+! INSTEAD\@?

    \item [Build regexes]  The words are reassembled to produce a regex
        matching the log line, and a counter is maintained for each
        regex.  Contiguous sequences of \texttt{.+} in the newly
        reassembled regexes are collapsed to a single \texttt{.+}; any
        resulting duplicate regexes are combined, and their counters
        added together.  If a regex's counter is lower than the
        threshold supplied by the user the regex is discarded; this
        second threshold is independent of the threshold used to
        differentiate between fixed and variable words, but once again 1\%
        of log lines is a good starting point.  The new regexes are
        displayed for the user to add to the database, either as new rules
        or merged into the regexes of existing rules; the counter for
        each regex is also displayed, giving the user an indication of
        how many of the log lines that regex should match.  Discarding
        regexes will result in some of the log lines not being matched;
        when the rules have been augmented with the new regexes, the
        original log files should be parsed again, and any remaining
        unparsed log lines used as input to this utility.

@@

A second utility is also provided that reads a list of new regexes and
the input given to the first utility.  It tries to match each input log
line against each regex, counting the number of log lines that match
each regex, warning the user if an input log line is matched by more
than one regex, and additionally warning if an input log line is not
matched by any regex.  It displays a summary of how many input log lines
each regex matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the regexes
produced by the first utility are correctly matching the input log lines
they are based upon.  A future version of this utility will also group
input log lines by regex, so the user can tweak the regexes if
required.

These utilities are not intended to create perfect regexes, but they
greatly reduce the effort required to parse new or different log lines.

@@

The rules and actions exhibit emergent behaviour that is far more
complicated than the behaviour explicitly encoded in the algorithm.  The
top-level parser could be written in pseudo-code like so (with indentation
level indicating flow control):

@@

    \item [matches] The fields in the line captured by the rule's regex.

@@

    \item [SAVE\_DATA] Find the correct mail based on the queueid in
        the log line, and save the data extracted by the regex to it.

@@

    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
        signal; the active connection for
        that \daemon{smtpd} must be cleaned up, entered in the database,
        and deleted from the state tables.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations --- after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur
        in the intervening hours: DNS timeouts, timeouts reading data
        from the client, etc\@.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

@@

This section presented the core of the parser, starting with a very high
level view and the initial complications that arose.  A flow chart showing
the paths a mail may take through the nascent simplified algorithm was
provided, followed by an explanation of those paths, and a discussion of
the parser's emergent behaviour --- the data from the log files creates the
paths in the flow chart, they are not specified anywhere in the parser.
The framework which holds the parser together was covered next, after which
came a description of the current actions provided by the parser, and the
algorithm for analysing unparsed log lines to create regexes for new
rules.  Detecting, diagnosing and defeating complications forms the largest
single portion of this section, mirroring the development of the parser.
The complications are described in the order they were overcome, with
subsequent problems affecting fewer mails (often by an order or magnitude),
though the time required to solve problems increased with each successive
problem.
